\titleNameLong{}

### 2015-12-08 FinalProject/Tex/paper/paper.tex
fixed various grammatical and line break issues

\begin{abstract}
This article is a survey of the current state of general-purpose computing on
GPUs (GPGPUs) and the use of heterogeneous applications on them. The underlying
architecture of GPUs is well-suited to problems with a high level of data
parallelism and has traditionally been used for rendering computer graphics.
Performing general-purpose computing on GPUs is a relatively recent development,
having only become practical and popular in the past decade or so. A brief
overview of a typical GPU architecture is presented, and the challenges of
applying this architecture to heterogeneous problems are discussed. Several
different current areas of research are explored, dynamic warp scheduling and
hardware tuning, reducing memory contention, and software solutions to the
challenges of GPGPU. While the challenges presented are difficult to overcome,
general-purpose computing on GPUs continues to make progress in the areas of
performance, energy, and power.
\end{abstract}

@@

In a way, the problems of GPGPU are an extension of those sometimes seen in a
traditional desktop multiprocessing system. Even when under a heavy compute
load, it is not always easy to keep all of the cores doing useful work in any
given cycle. Between limitations in structural units, dependences between
instructions, and latencies of memory accesses, the processor cannot help but
stall sometimes. This problem is exacerbated by orders of magnitude when solving
heterogeneous problems on a GPU. GPUs were originally designed to operate on
graphics processing, problems that are well-structured and have simple, inherent
data parallelism. Problems that do not contain parallelism that is easily mapped
to the architecture of a GPU fail to keep all of the cores in a GPU busy or
strain the memory bandwidth past the GPU's ability to hide latency.

@@

\section{Architectural Overview} \label{sec:architecture}
\subsection{Streaming Multiprocessors}
Modern unified GPUs consist of multiple streaming multiprocessors (SMs). While
there are many GPU architectures available today, the following description is
based off of NVIDIA's Maxwell architecture, which was released in 2014. These
SMs are similar to SIMD processors and typically contain the following
components:
\begin{description}
  \setlength\itemsep{0.5em}
  \item[A large register file] The register file is shared across all threads
  and typically contains thousands of registers, many more than are usually
  found in a CPU. NVIDIA's Maxwell architecture has 64K 32-bit registers in each
  SM.
  \item[Multiple caches] Maxwell includes a unified L1/texture cache, which used
  as a coalescing buffer for memory accesses. The buffer gathers requested data
  and sends the entire group to a warp. There are also additional caches TODO.
  \item[Shared memory] There are 96KB available per SM in Maxwell. However,
  thread blocks are limited to 48KB each. Shared memory for a thread-block is
  only accessible by threads within the same thread-block.
  \item[Multiple warp schedulers] Warp schedulers dispatch and allocate
  resources. Maxwell contains four schedulers per SM, and functional units are
  assigned to a particular scheduler. Using a power-of-two number of CUDA cores
  per SM allows for simple partitioning among schedulers. One ready warp is
  scheduled per cycle, which allows the GPU to hide the latency of stalled
  warps.
\end{description}

@@

Grids are what actually run on the GPU, and they are composed of multiple
``thread- blocks'', also called a cooperative thread array (CTA) outside of
NVIDIA's nomenclature. Maxwell supports up to 64K blocks within a grid. Thread-
blocks contain potentially many threads, on the order of 1024 for modern
architectures. Each thread within a thread-block has the same life-cycle as all
the others, meaning that they are dispatched to and swapped from a SM at the
same time.

Individual threads exist within a thread-block. These threads are grouped into
``warps'', usually around 32 at a time. The warp is the fundamental unit of
execution on GPUs. Threads within a warp are mapped to a vector functional unit
such that all threads execute the same instruction on different data, just a
like a SIMD unit. With this in mind, it would be more accurate to simply call
``threads'' a sequence of SIMD lane instructions, but this survey will continue
to refer to them as threads to maintain consistency with the literature. The
term ``Warp'' is the term used by NVIDIA, while AMD and OpenCL use
``wavefront''. Maxwell supports up to 64 warps resident per SM. Note that this
does not mean that all 64 warps are executing at the same time, but that the
architecture can support that many warp contexts concurrently. The SM maintains
a warp pool to store the context of all running threads. Warps are scheduled by
the warp schedulers from this pool.

@@

\section{Dynamically-Launched Helpers} \label{sec:dynamic}
\subsection{Assist warps}
As we have seen, GPUs are capable of supporting thousands of threads, but
bottlenecks in heterogeneous applications limit performance. \cite{AssistWarps}
present the idea of using ``assist warps'' in their Core-Assisted Bottleneck
Acceleration (CABA) scheme. These special warps are generated by dedicated
hardware and are used as helper threads to increase performance and efficiency.
Assist warps share the same context as the parent warp and uses some of the
available registers, since investigation shows that 24\% of the register file is
unallocated on average.

One primary use for assist warps is memory compression. Instead of using
dedicated hardware for compressing and decompressing data from memory, this
approach uses existing hardware that is not being utilized. Another advantage of
not using dedicated hardware is the ability to change compression algorithms
flexibly, since different applications have varying access patterns that are
suited for different compression algorithms. Another use for assist warps is in
memoization, in which frequently- computed values from main memory are cached
and used as lookup tables, reducing the number of loads and stores required. One
final use suggested by the authors is using an idle memory pipeline to prefetch
values during computation to reduce the memory pressure.

CABA contains a mix of hardware and software. While software-only or hardware-
only options are possible, the former induces high overhead and complicates
communication between primary and helper warps, and the latter makes having a
flexible interface more difficult. The combination suggested by the authors
dynamically inserts instructions into the execution stream in the form of assist
warps. This requires stores of commonly used assist warps, a controller to
trigger and manage assist warp execution, and an assist warp buffer inside the
instruction buffer that holds triggered warps until they are scheduled. Helper
threads share context with the parent warp, and register values in helper
threads are not preserved between invocations, reducing the overhead of this
technique and does not require any additional resources.

@@

\subsection{Dynamic Thread Block Launch}
The authors of Dynamic Thread Block Launch (DTBL) suggest a somewhat similar
approach, this time launching thread blocks dynamically instead of assist warps
\cite{DynamicThreadBlockLaunch}. The motivation behind this technique is that
while emerging general-purpose applications do not have simple, regular
structure that is easy to exploit at the highest level, they still do contain
``pockets'' of parallelism at smaller levels. The authors explain that these
pockets can be exploited with existing architecture using device-side nested
kernel launching, but at a high cost. For example, CUDA supports parent kernels
launching child kernels through device-side API calls. These child kernels can
start at any time after being launched by their parent. This method does exploit
pockets of parallelism, but at a high cost. On average, this CUDA method results
in a 36.1\% decrease in performance, and also consumes significant amounts of
memory by storing pending child kernels.

The problems with existing implementations boil down to four characteristics:
\begin{description}
  \setlength\itemsep{0.5em}
  \item [High kernel density] Kernel launches are expensive, and typical
  applications currently require many kernels. For example, a typical breadth-
  first search requires about 3,000 device kernel launches, resulting in a high
  overhead and memory footprint.
  \item [Low compute intensity] While many kernels are launched using these
  methods, they are usually fine-grained and do not contain high amounts of
  parallelism. The average number of threads in these kernels is around 40, which
  is close to the warp size and could be better served by a more fine-grained
  approach.
  \item [Workload similarity] The operations performed by each dynamically
  launched kernel are usually similar, but contain different levels of parallelism
  due to differing configurations or data.
  \item [Low concurrency/scheduling efficiency] Due to the low compute intensity
  of the device kernels, multiple kernels should be able to execute in parallel.
  However, GPU architectures place limits on the number of different kernels that
  can execute concurrently, and since each kernel has a relatively small number of
  warps within it, the SMs are underutilized and struggle to hide memory
  latencies.
\end{description}

@@

\section{Dynamic Scheduling} \label{sec:scheduling}
\subsection{Critical Warps}
GPUs manage to hide latency by swapping out stalled warps, but current
performance is not good enough. This latency hiding is limited by the following
characteristics of heterogeneous applications: imbalanced workloads in parallel
warps, diverging branch behavior, irregular memory access patterns, and shared
cache contention \cite{CoordinatedWarpScheduling}. These imbalances and
irregularities result in differences in execution time of parallel warps. The
slowest warp in a group is called the ``critical'' Since warps are blocked for
synchronization purposes until all of them are finished, this needlessly extends
the life of all the other warps and limits the performance of the application as
a whole. The authors of Criticality-aware warp acceleration (CAWA) demonstrate
that the average difference in execution time between the fastest and slowest
warp is 45\% and can reach as high as 70\%.

@@

\subsection{Variable Warp Size}
While some of the past sections have discussed added new execution objects to
perform work, this technique pertains to splitting and re-merging existing
warps. There are two types of divergence in GPU execution: control flow
divergence and memory divergence. The former deals with complex control flow in
large warps resulting in performance loss, while the latter is related to how
well memory requests are coalesced into single warps. When warps are too small,
horizontal locality of memory accesses is lost; this effect is termed ``memory
convergence slip.''

In addition, workloads can be classified into three groups depending on how
their performance is affected by warp size. Divergent warps perform poorly as
warp size increases, due to control flow divergence as explained above. Warp-
size insensitive warps see little to no performance changes with warp size.
Convergent warps actually see better performance as warp size increases, due to
coalescing execution and memory accesses into a single warp. Increasing warp
size is better for the latter two types, since larger warps have better energy
efficiency.

Recent applications are increasingly divergent
\cite{VariableWarpSize}: raytracing, molecular dynamics simulations, game
physics simulations, and graph processing are all common applications for GPGPU
and all contain many divergent warps. To combat this, warp sizes can be adjusted
dynamically. A minimum of 4 is a good floor value, since graphics workloads
commonly process ``quads'' (threads in groups of 4) and there are diminishing
returns for warp sizes any smaller than this. This technique works by splitting
the traditional datapath into 8 slices, where each slice can fetch, decode, and
issue independently. Slices are statically assigned threads in a linear fashion
(0-3 to slice 0, 4-7 to slice 1, etc.), and threads cannot move between slices.

@@

\subsection{Hardware Worklists}
There are two common approaches to mapping graph algorithms to GPUs. The first
is topology-driven, in which the current thread index is used to determine on
which nodes to operate. This approach launches a new kernel for every iteration
and visits all nodes. Since launching kernels is expensive and some or many
nodes do not have any useful work to do in a given iteration, this results in
inefficiencies. The other approach is data-driven, where shared software
worklists (SWWL) are used to determine which nodes have useful work for a given
iteration. Only one kernel is launched at the start, and it is refilled with
work as necessary each iteration. This approach avoids all of the useless work
found in the topology-driven method, but can require a very large worklist,
which makes it prone to memory contention or access irregularity. Because of
this drawback, data-driven mapping is impractical unless aggressive software
optimizations are made.

@@

Even with such optimizations, the data-driven approach is not ideal due to
remaining memory contention on pushes and poor load balancing. Using fine-
grained hardware worklists (HWWL) instead of software worklists is one method of
dealing with such drawbacks
\cite{UsingFineGrainHardwareWorklists}. This works by exposing the HWWL to the
software through simple additions to the ISA. This way an application utilizing
a SWWL can be easily adapted to use a HWWL. The microarchitecture of the GPU is
modified to support a HWWL, distributing work to each lane in the GPU. To enable
better load balancing, empty banks pull from other banks instead of exiting.
There are two policies for when to redistribute work: interval-based and on-
demand. The former periodically redistributes work at a given sampling interval.
This has the advantage of better load balancing, but a higher cost. The
alternative, on- demand redistribution, only redistributes when a bank pulls or
pushes. This saves energy but has poorer load balancing than the interval-based
policy.

@@

\subsection{Dynamic Frequency and Parallelism Adjustment}
The last technique in this section deals with dynamically monitoring resource
requirements and adjusting the SM frequency, memory frequency, and number of
threads executing. A majority of GPU applications are bottlenecked by either a
lack of compute resources, memory bandwidth, or a limited data cache, and so
executing the maximum number of threads and the highest frequencies is not
always beneficial to performance. If a given resource is bottle-necked, the
other resources become underutilized. For example, the NVIDIA Fermi architecture
has 48KB of data cache. When spread out over the maximum number of threads that
can execute at the same time, this leaves only 30 bytes for each.

If the characteristics of an application's workload were known before execution,
the GPU could be tuned to execute the optimal number of threads at the optimal
SM and DRAM frequencies. Unfortunately, this is often not possible: workloads
are heavily dependent on the given application, its input, and the specific GPU.
Equalizer \cite{EqualizerTuningOfResources} is a runtime system that runs on
individual SMs, determines the resource requirements of kernels, and adjusts the
number of executing threads and frequencies of the SM and memory.

@@

To linearize the control flow, loops are analyzed with structural analysis.
After the analysis has determined control flow, the direction of the edges in
the CFG are reversed, and the CFG is processed in a depth-first post- order
traversal. This yields a valid schedule.

Applying these algorithms results in code that is 2.7\% slower on average, with
little to no area, power, or energy implications. The main motivation for this
approach is that is allows for less hardware design complexity and verification
costs. From the programmer's point of view, there is no difference. In addition,
using divergence stack control sometimes results in nondeterministic order of
execution. Using the variance analysis in this approach allows the compiler to
make stronger guarantees at compile time.

@@

\section{Other Research} \label{sec:memory}
\subsection{Warp Data Compression}
To increase the breadth of this survey, two additional topics are discussed that
do not fit well in the categories so far. The first is compression and
decompression of register file accesses on a warp level. GPU register files
contribute up to 15\% of the total GPU chip power, so there are significant
savings available if the register file can be made more power efficient. One
past approach has been turning off unused parts of the register file. This saves
both leakage and dynamic power. Another approach takes advantage of register use
locality with register file caches.

@@

To compress the values used in warps, value similarity is key. This is the
arithmetic distance between register values. Values across threads in a warp are
typically similar, since data arrays are commonly accessed using the thread IDs
of successive threads, and the dynamic range of input data is small for many
things. Power can be saved by avoiding the repeated storing, reading, and
writing of redundant data. Distances between successive thread registers are
calculated and placed into one of four bins: zero distance, which means the two
values are the same; 128 bin, where the values differ by at most 128; 32K bin,
where they differ by at most 32K, and the random bin, which holds all values
that do not fall into any of the first three.

Keeping in mind the trade-off of compression ratio vs latency, the base-delta-
immediate (BDI) scheme is most suitable warp compression. This scheme uses a
base value, and subsequent deltas are added to the base to determine each
subsequent value. Computed values are compressed before writing to the register
file and decompressed before reading. Since most warp instructions have two
sources and one destination, it makes sense to have two decompression units and
a single compression unit to process one warp per cycle. Since modern GPU
architectures can process two warps per cycle, the number of each of these is
doubled, for a total of four decompression units and two compression units.

@@

\subsection{Memory Contention in Heterogeneous Multiprocessing Systems}
Modern computing system architectures include both a CPU and GPU in a master-
slave configuration, in which the CPU offloads heavy parallel computations to
the GPU. In the future, there will likely be many CPUs and GPUs operating in
conjunction within a single system. These network-on-chip (NoC) configurations
are called heterogeneous multiprocessing architectures
\cite{ManagingConcurrencyInHeterogeneous}.

CPU applications tend to be more latency sensitive, while GPU applications are
more bandwidth sensitive. In such heterogeneous architectures containing both,
GPUs tend to cannibalize the performance of CPUs much more than the other way
around. The goal of the technique described here is to prevent this
cannibalization and improve the performance of both kinds of computation. This
technique assumes a tile-based mesh architecture consisting of both CPUs and
GPUs. The CPU to GPU ratio is 2:1 because a single GPU SM occupies roughly half
the space of a CPU core. The last level cache of the system is shared by CPUs
and GPUs, and they use a shared network and shared memory channels.

GPU concurrency can have a positive or negative effect on other GPU performance.
Increasing the number of concurrent warps can help performance but can also
thrash the cache or cause memory congestion. The effects of GPU concurrency on
CPU performance, however, are much more negative. Memory congestion hurts CPUs
much more than GPUs, as they do not have enough thread-level parallelism to hide
the latency.

@@

% Acknowledgments
\begin{acks}
The author would like to thank Dr. Penry for teaching this course and for
providing extensive design and learning opportunities in this course. It has
been a lot of work, but this course has helped the author know that he has
chosen the right field.
\end{acks}

>>>>

\begin{abstract}
This article is a survey of the current state of general-purpose computing on
GPUs (GPGPUs) and the use of heterogeneous applications on them. The underlying
architecture of GPUs is well-suited to problems with a high level of data
parallelism and has traditionally been used for rendering computer graphics.
Performing general-purpose computing on GPUs is a relatively recent development,
having only become practical and popular in the past decade or so. A brief
overview of a typical GPU architecture is presented, and the challenges of
applying this architecture to heterogeneous problems are discussed. Several
different current areas of research are explored, including dynamic warp
scheduling and hardware tuning, reducing memory contention, and software
solutions to the challenges of GPGPU. While the challenges presented are
difficult to overcome, general-purpose computing on GPUs continues to make
progress in the areas of performance, energy, and power.
\end{abstract}

@@

In a way, the problems of GPGPU are an extension of those sometimes seen in a
traditional desktop multiprocessing system. Even when under a heavy compute
load, it is not easy to keep all of the cores doing useful work in any given
cycle. Between limitations in structural units, dependences between
instructions, and latencies of memory accesses, the processor cannot help but
stall sometimes. This problem is exacerbated by orders of magnitude when solving
heterogeneous problems on a GPU. GPUs were originally designed to operate on
graphics processing, problems that are well-structured and have simple, inherent
data parallelism. Problems that do not contain parallelism that is easily mapped
to the architecture of a GPU fail to keep all of the cores in a GPU busy or
strain the memory bandwidth past the GPU's ability to hide latency.

@@

\section{Architectural Overview} \label{sec:architecture}
\subsection{Streaming Multiprocessors}
Modern unified GPUs consist of multiple streaming multiprocessors (SMs). Figure
\ref{fig:baseline} shows a baseline GPU configuration
\cite{DynamicThreadBlockLaunch}. While there are many GPU architectures
available today, the following description is based off of NVIDIA's Maxwell
architecture, which was released in 2014. These SMs are similar to SIMD
processors and typically contain the following components:
\begin{description}
  \setlength\itemsep{0.5em}
  \item[A large register file] The register file is shared across all threads
  and typically contains thousands of registers, many more than are usually
  found in a CPU. NVIDIA's Maxwell architecture has 64K 32-bit registers in each
  SM.
  \item[Multiple caches] Maxwell includes a unified L1/texture cache, which used
  as a coalescing buffer for memory accesses. The buffer gathers requested data
  and sends the entire group to a warp. There are also additional caches TODO.
  \item[Shared memory] There are 96KB available per SM in Maxwell. However,
  thread blocks are limited to 48KB each. Shared memory for a thread-block is
  only accessible by threads within the same thread-block.
  \item[Multiple warp schedulers] Warp schedulers dispatch and allocate
  resources. Maxwell contains four schedulers per SM, and functional units are
  assigned to a particular scheduler. Using a power-of-two number of CUDA cores
  per SM allows for simple partitioning among schedulers. One ready warp is
  scheduled per cycle, which allows the GPU to hide the latency of stalled
  warps.
\end{description}

@@

Grids are what actually run on the GPU, and they are composed of multiple
``thread-blocks'', also called a cooperative thread array (CTA) outside of
NVIDIA's nomenclature. Maxwell supports up to 64K blocks within a grid.
Thread-blocks contain potentially many threads, on the order of 1024 for modern
architectures. Each thread within a thread-block has the same life-cycle as all
the others, meaning that they are dispatched to and swapped from a SM at the
same time.

Individual threads exist within a thread-block. These threads are grouped into
``warps'', usually around 32 at a time. The warp is the fundamental unit of
execution on GPUs. Threads within a warp are mapped to a vector functional unit
such that all threads execute the same instruction on different data, just a
like a SIMD unit. With this in mind, it would be more accurate to call
``threads'' a sequence of SIMD lane instructions, but this survey will continue
to refer to them as threads to maintain consistency with the literature. The
term ``warp'' is the term used by NVIDIA, while AMD and OpenCL use
``wavefront''. Maxwell supports up to 64 warps resident per SM. Note that this
does not mean that all 64 warps are executing at the same time, but that the
architecture can support that many warp contexts concurrently. The SM maintains
a warp pool to store the context of all running threads. Warps are scheduled by
the warp schedulers from this pool.

\begin{figure}
  \centering
  \includegraphics[]{BaselineGPU.png} \caption{Test caption} \label{fig:baseline}
\end{figure}

@@

\section{Dynamically-Launched Helpers} \label{sec:dynamic}
\subsection{Assist warps}
As we have seen, GPUs are capable of supporting thousands of threads, but
bottlenecks in heterogeneous applications limit performance. Core-Assisted
Bottleneck Acceleration \cite{AssistWarps} is a scheme that uses ``assist
warps.'' These special warps are generated by dedicated hardware and are used as
helper threads to increase performance and efficiency. Assist warps share the
same context as the parent warp and uses some of the available registers, since
investigation shows that 24\% of the register file is unallocated on average.

One primary use for assist warps is memory compression. Instead of using
dedicated hardware for compressing and decompressing data from memory, this
approach uses existing hardware that is not being utilized. Another advantage of
not using dedicated hardware is the ability to change compression algorithms
flexibly, since different applications have varying access patterns that are
suited for different compression algorithms. Assist warps can also be used for
memoization, in which frequently-computed values from main memory are cached and
used as lookup tables, reducing the number of loads and stores required. One
final use suggested by the authors is using an idle memory pipeline to prefetch
values during computation to reduce the memory pressure.

CABA contains a mix of hardware and software. While software-only or
hardware-only options are possible, the former induces high overhead and
complicates communication between primary and helper warps, and the latter makes
having a flexible interface more difficult. The combination suggested by the
authors dynamically inserts instructions into the execution stream in the form
of assist warps. This requires stores of commonly used assist warps, a
controller to trigger and manage assist warp execution, and an assist warp
buffer inside the instruction buffer that holds triggered warps until they are
scheduled. Helper threads share context with the parent warp, and register
values in helper threads are not preserved between invocations, reducing the
overhead of this technique and does not require any additional resources.

@@

\subsection{Dynamic Thread Block Launch}
The authors of Dynamic Thread Block Launch (DTBL) suggest a somewhat similar
approach, this time launching thread blocks dynamically instead of assist warps
\cite{DynamicThreadBlockLaunch}. The motivation behind this technique is that
while emerging general-purpose applications do not have a simple, regular
structure that is easy to exploit at the highest level, they still do contain
``pockets'' of parallelism at smaller levels. The authors explain that these
pockets can be exploited with existing architecture using device-side nested
kernel launching, but at a high cost. For example, CUDA supports parent kernels
launching child kernels through device-side API calls. These child kernels can
start at any time after being launched by their parent. This method does exploit
pockets of parallelism, but this CUDA method results in an average of 36.1\%
decrease in performance, and also consumes significant amounts of memory by
storing pending child kernels.

The problems with existing implementations reduce down to four characteristics:
\begin{description}
  \setlength\itemsep{0.5em}
  \item [High kernel density] Kernel launches are expensive, and typical
  applications currently require many kernels. For example, a typical
  breadth-first search requires about 3,000 device kernel launches, resulting in
  a high overhead and memory footprint.
  \item [Low compute intensity] While many kernels are launched using these
  methods, they are usually fine-grained and do not contain high amounts of
  parallelism. The average number of threads in these kernels is around 40, which
  is close to the warp size and could be better served by a more fine-grained
  approach.
  \item [Workload similarity] The operations performed by each dynamically
  launched kernel are usually similar, but contain different levels of parallelism
  due to differing configurations or data.
  \item [Low concurrency/scheduling efficiency] Due to the low compute intensity
  of the device kernels, multiple kernels should be able to execute in parallel.
  However, GPU architectures place limits on the number of different kernels that
  can execute concurrently, and since each kernel has a relatively small number of
  warps within it, the SMs are underutilized and struggle to hide memory
  latencies.
\end{description}

@@

\section{Dynamic Scheduling} \label{sec:scheduling}
\subsection{Critical Warps}
GPUs manage to hide latency by swapping out stalled warps, but current
performance is not good enough. This latency hiding is limited by the following
characteristics of heterogeneous applications: imbalanced workloads in parallel
warps, diverging branch behavior, irregular memory access patterns, and shared
cache contention \cite{CoordinatedWarpScheduling}. These imbalances and
irregularities result in differences in execution time of parallel warps. The
slowest warp in a group is called the ``critical'' warp. Since warps are blocked
for synchronization purposes until all of them are finished, this needlessly
extends the life of all the other warps and limits the performance of the
application as a whole. The authors of Criticality-aware warp acceleration
(CAWA) demonstrate that the average difference in execution time between the
fastest and slowest warp is 45\% and can reach as high as 70\%.

@@

\subsection{Variable Warp Size}
While some of the past sections have discussed added new execution objects to
perform work, this technique pertains to splitting and re-merging existing warps
\cite{VariableWarpSize}. There are two types of divergence in GPU execution:
control flow divergence and memory divergence. The former deals with complex
control flow in large warps resulting in performance loss, while the latter is
related to how well memory requests are coalesced into single warps. When warps
are too small, horizontal locality of memory accesses is lost; this effect is
termed ``memory convergence slip.''

In addition, workloads can be classified into three groups depending on how
their performance is affected by warp size. Divergent warps perform poorly as
warp size increases, due to control flow divergence as explained above.
Warp-size insensitive warps see little to no performance changes with warp size.
Convergent warps actually see better performance as warp size increases, due to
coalescing execution and memory accesses into a single warp. Increasing warp
size is better for the latter two types, since larger warps have better energy
efficiency.

Recent applications are increasingly divergent: raytracing, molecular dynamics
simulations, game physics simulations, and graph processing are all common
applications for GPGPU and all contain many divergent warps. To combat this,
warp sizes can be adjusted dynamically. A minimum of 4 is a good floor value,
since graphics workloads commonly process ``quads'' (threads in groups of 4) and
there are diminishing returns for warp sizes any smaller than this. This
technique works by splitting the traditional datapath into 8 slices, where each
slice can fetch, decode, and issue independently. Slices are statically assigned
threads in a linear fashion (0-3 to slice 0, 4-7 to slice 1, etc.), and threads
cannot move between slices.

@@

\subsection{Hardware Worklists}
There are two common approaches to mapping graph algorithms to GPUs
\cite{UsingFineGrainHardwareWorklists}. The first is topology-driven, in which
the current thread index is used to determine on which nodes to operate. This
approach launches a new kernel for every iteration and visits all nodes. Since
launching kernels is expensive and some or many nodes do not have any useful
work to do in a given iteration, this results in inefficiencies. The other
approach is data-driven, where shared software worklists (SWWL) are used to
determine which nodes have useful work for a given iteration. Only one kernel is
launched at the start, and it is refilled with work as necessary each iteration.
This approach avoids all of the useless work found in the topology-driven
method, but can require a very large worklist, which makes it prone to memory
contention or access irregularity. Because of this drawback, data-driven mapping
is impractical unless aggressive software optimizations are made.

@@

Even with such optimizations, the data-driven approach is not ideal due to
remaining memory contention on pushes and poor load balancing. Using
fine-grained hardware worklists (HWWL) instead of software worklists is one
method of dealing with such drawbacks. This works by exposing the HWWL to the
software through simple additions to the ISA. This way an application utilizing
a SWWL can be easily adapted to use a HWWL. The microarchitecture of the GPU is
modified to support a HWWL, distributing work to each lane in the GPU. To enable
better load balancing, empty banks pull from other banks instead of exiting.
There are two policies for when to redistribute work: interval-based and
on-demand. The former periodically redistributes work at a given sampling
interval.This has the advantage of better load balancing, but a higher cost. The
alternative, on-demand redistribution, only redistributes when a bank pulls or
pushes. This saves energy but has poorer load balancing than the interval-based
policy.

@@

\subsection{Dynamic Frequency and Parallelism Adjustment}
The last technique in this section deals with dynamically monitoring resource
requirements and adjusting the SM frequency, memory frequency, and number of
threads executing. A majority of GPU applications are bottlenecked by either a
lack of compute resources, memory bandwidth, or a limited data cache, and so
executing the maximum number of threads and the highest frequencies is not
always beneficial to performance \cite{EqualizerTuningOfResources}. If a given
resource is bottle-necked, the other resources become underutilized. For
example, the NVIDIA Fermi architecture has 48KB of data cache. When spread out
over the maximum number of threads that can execute at the same time, this
leaves only 30 bytes for each.

If the characteristics of an application's workload were known before execution,
the GPU could be tuned to execute the optimal number of threads at the optimal
SM and DRAM frequencies. Unfortunately, this is often not possible: workloads
are heavily dependent on the given application, its input, and the specific GPU.
Equalizer is a runtime system that runs on individual SMs, determines the
resource requirements of kernels, and adjusts the number of executing threads
and frequencies of the SM and memory.

@@

To linearize the control flow, loops are analyzed with structural analysis.
After the analysis has determined control flow, the direction of the edges in
the CFG are reversed, and the CFG is processed in a depth-first post-order
traversal. This yields a valid schedule.

Applying these algorithms results in code that is 2.7\% slower on average, with
little to no effect on area, power, or energy. The main motivation for this
approach is that is allows for less hardware design complexity and verification
costs. From the programmer's point of view, there is no difference. In addition,
using divergence stack control sometimes results in nondeterministic order of
execution. Using the variance analysis in this approach allows the compiler to
make stronger guarantees at compile time.

@@

\section{Other Research} \label{sec:memory}
\subsection{Warp Data Compression}
To increase the breadth of this survey, two additional topics are discussed that
do not fit well in the categories so far. The first is compression and
decompression of register file accesses on a warp level. GPU register files
contribute up to 15\% of the total GPU chip power, so there are significant
savings available if the register file can be made more power efficient. One
past approach has been turning off unused parts of the register file, which
saves both leakage and dynamic power. Another approach takes advantage of
register use locality with register file caches.

@@

To compress the values used in warps, the arithmetic distance between register
values should be small. This is known as value similarity. Values across threads
in a warp are typically similar, since data arrays are commonly accessed using
the thread IDs of successive threads, and the dynamic range of input data is
often small as well. Power can be saved by avoiding the repeated reading and
writing of mostly redundant data. Distances between successive thread registers
are thus calculated and used to place the registers into one of four bins: zero
distance, which means the base register and current register are the same; 128
bin, where the values differ by at most 128; 32K bin, where they differ by at
most 32K, and the random bin, which holds all values that do not fall into any
of the first three.

Keeping in mind the trade-off of compression ratio vs latency, the
base-delta-immediate (BDI) scheme is most suitable for warp compression. This
scheme uses one register a base value, and subsequent deltas are added to the
base to determine each subsequent register's value. Computed values are
compressed before writing to the register file and decompressed before reading.
Since most warp instructions have two sources and one destination, it makes
sense to have two decompression units and a single compression unit to process
one warp per cycle. Since modern GPU architectures can process two warps per
cycle, the number of each of these is doubled, for a total of four decompression
units and two compression units.

@@

\subsection{Memory Contention in Heterogeneous Multiprocessing Systems}
Modern computing system architectures include both a CPU and GPU in a
master-slave configuration, in which the CPU offloads heavy parallel
computations to the GPU. In the future, there will likely be many CPUs and GPUs
operating in conjunction within a single system. These network-on-chip (NoC)
configurations are called heterogeneous multiprocessing architectures
\cite{ManagingConcurrencyInHeterogeneous}.

CPU applications tend to be more latency sensitive, while GPU applications are
more bandwidth sensitive. In such heterogeneous architectures containing both
kinds of processors, GPUs tend to cannibalize the performance of CPUs much more
than the other way around. The goal of the technique described here is to
prevent this cannibalization and improve the performance of both kinds of
computation. This technique assumes a tile-based mesh architecture consisting of
both CPUs and GPUs. The CPU to GPU ratio is assumed to be 2:1 because a single
GPU SM occupies roughly half the space of a CPU core. The last level cache of
the system is shared by CPUs and GPUs, and they use a shared network and shared
memory channels.

GPU concurrency can have a positive or negative effect on other GPU performance,
since increasing the number of concurrent warps can help performance but can
also cause cache thrashing and/or memory congestion. The effects of GPU
concurrency on CPU performance, however, are much more negative. Memory
congestion hurts CPUs much more than GPUs, as they do not have enough
thread-level parallelism to hide the latency.

@@

% Acknowledgments
\begin{acks}
The author would like to thank Dr. Penry for teaching this course and for
providing extensive design and learning opportunities throughout the semester.
It has been a lot of work, but this course has given the author a glimpse into
what computer architecture is all about and helped him know that he has chosen
the right field.
\end{acks}
